\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction} 

Several years ago, when I was programming, I was fascinated
by the concept of meta-programming, i.e., generating code with code
so that I can write less code. Generally, you write some code for
a real problem. Even if your target is to generate code, you have
to execute the code to get the result. This sounds nonsense to a 
mathematician, but this is actually how we handle mathematical 
logic, i.e., we define our language as mathematical objects 
(We define the well formed formula or proof as sets) 
just as we define a group with set theory. When we want to prove
something, we follow the predefined syntax (e.g. a deduction system)
constructing a proof to consolidate a belief. We never infer anything
about this syntax itself once it is defined. This is because if we
want to infer about the inference itself, we may end up with the 
Rousell's paradox. However, we study the structure of reasoning
safely as a mathematical object. Do we still infer about the inference
so that it might trigger the paradox? The basic observation is that 
we treat it as a {\bf standalone} object, an identical but distinct 
object from the meta one we use. This is also adopted by 
computer scientists, i.e., they can write a program (usually called 
a compiler) that turns or generates the source code (as objects) 
written in the same or different language to a program. Typically, 
the language they use to write the program is called the meta language, 
while the outcome language is the target language. They are generally 
independent. One is analysized as objects of the other. 

But why do I have to emphasize programming language? A shocking fact is 
that defining inference rule and applying it to another inference is 
widely and naturally accepted in programming. For example, you may have
proven that to prove $\phi$ or $\psi$, you only need to prove $\psi$ 
assuming $\phi$ is wrong. I used this in the proof of 
{\it The boundary of a union is contained in the union of boundaries} 
($\partial(A\cup B)\subseteq \partial A \cup\partial B$). 
When proving that proposition in {\it Topology Theory}, I can never 
write a sentence like $\forall A,B\in Props,\phi\vee\psi \leftrightarrow
\neg\phi\rightarrow\psi$, because the Topology Theory knows nothing
about the concept $Props$. The proof of it is outside the theory,
i.e., in the second-order language, but in our language, we can
prove both. When it comes to the programming language, you can also
prove them in the same context. According to the Curry-Howard 
Correspondence \cite{Curry-Howard,Curry-Howard-Scheme}, such a
proposition $(\neg\phi\to\psi)\to\phi\vee\psi$ is understood as a 
function, which, given sets \footnote{For an intuitive illustration, 
here I use set, but it later turns out a type.} $\phi$ and $\psi$, 
gives you a function mapping any function from 
$\neg\phi\to\psi$ to the coproduct $\phi\vee\psi$. (This is 
called a dependent type, showing a kind of quantification. See 
\autoref{type-theory} for a formal description.) When using it to
prove the propsition about boundary($\partial$), you apply this
function to some $\phi$, $\psi$ and $\neg\phi\to\psi$ to generate
a $\phi\vee\psi$. Certainly, this is not a set-theoretic function,
since $\phi$ and $\psi$ traverse the whole universe of sets. Actually,
the behavior is like the $\forall$ quantification, which does traverse
all sets. When applying it to some element, the compiler will determine
the suitable set $\phi$ and $\psi$, which won't cause you any trouble
except that you want to apply it to itself, i.e., one can still lift
the order of the language or he/she gets the paradox. A clever solution 
towards this is building a hierarchical universe as von Neumann did 
\cite{von-Neumann-universe}. Once $\phi$ and $\psi$ are picked, the 
compiler will infer the order of our language. 

In our real world, this is what mathematicians want, an any order
language or a semantic language. For example, if we take the 
first-order language to be the language of group theory $(\cdot,e)$, 
then the (structured) set theory (model theory) is the second order one.
A proof about group properties can be done with merely group theory,
but do you really use it alone? As in any textbook, the author always
takes a group as a set and apply set theory to prove, and properties
beyond group theory (like morhpisms) are also discussed with set
theory. For a third order example, logical structures are also proved
somewhere. Recall that you must have encountered some problem in
group theory or set theory where you want to use a tautology to turn 
it into a logically equivalent problem but you have to check whether 
this tautology is correct or not. Some may argue that they can certainly 
introduce all these tautologies
into the deduction system (via schemes\footnote{A collection of
axioms. Scheme is also the name of a programming language}), so you 
only have to deal with the first order one \footnote{Unlike the behavior 
discussed above, a C++ compiler only accepts such functions with all 
quantifications at the beginning like schemes, i.e. the quantifier 
prenex form and only universal quantifications are accepted, while 
Haskell has a language extension to allow existential quantifications 
so that you can put the quantification at the middle of a sentence.
Idris, another programming language, adopts the hierarchical universe
with orders of the language. The macro system of C is a totally 
separation from the target language as a meta language. They coincide
with our handling in logic.} (the set theory is also flattened with
group theory as both are first order), but as you see, in your real 
practice, you'd like to verify it first and then apply it. (This means
the inference rules should be finite.) Without always 
explicitly specifying the order of the language, one can ignore this 
problem to some extent. A clever computer proof-assistant system will 
help you with it, or one can just use his/her brain to tell the order. 
Once this order is determined, the rest is a plain set theory problem 
first-orderly. (Note also that the set theory used during the 
development of first order language is different from the one developed 
by first order language later in the fact that they are in different
universe with different orders). Another notable part is that the
properties of a group (functions and equality) coincide with those of 
set (definable with set theory). Thus we can represent a group with a
set. From this perspective, group theory is a domain specific language
\footnote{This concept is also borrowed from progromming language.
In real world, a programmer only focus on the domain of his/her task.
They want to represent a real world problem with turing machine. Idris
provides a syntax suger to freely define any domain specific gramma.} 
(DSL) represented by set theory, or you can call set theory the frontend 
\footnote{The direct interface we write down in our books or on paper. 
In programming world, Tensorflow or PyTorch use python as their frontend 
language.} language for group theory, i.e. every set operation (only the 
equality comparison and binary operation) is mapped to an operation in 
group. 

The first order language together 
with sets (models) as its semantics is representative but it is not good 
at self referring (You have to lift the order of the language and treat 
it separately). In programming, once the computation process (proofs) can 
be efficiently represented by data structures (objects like sets)\footnote{
    For example the {\bf lisp} language family based directly on lambda
    calculus uses a list-like form as its only syntax, thus making it
    convenient to generate code.
}, we can use the computational feature at current session to generate
code and then run the result. In logic, you may want to define a function
to generate proof, which means we don't prove inside the deduction system
but prove the existence of the construction of a proof outside the 
deduction system. As a substitution, type theory embraces these benefits. 
The type theory contains two inherent layers, a layer of object and
another of reduction. The latter coincides with the inference rule,
namely the Curry-Howard correspondence \cite{Curry-Howard,
intuitionistic-type-theory}. It's very easy to represent a proof as an
object (function types) and the {\it $\beta$-reduction} is the inference
rule {\it modus ponens}. In fact, it can also be used as a better language
for homotopy theory \cite{homotopy-type-theory} or topos theory
\cite{Introduction-to-higher-order-categorical-logic,categorical-logic-and-type-theory}

The computational explanation for logic also raises other problems.
In general we can not compare two proofs. One can always write down
redundant sentences amidst a proof, as is a redundant calculation
abstraction. To find the shortest form of a calculation is known as
the normalization problem in typed lambda calculus\cite{Curry-Howard}.
Again, the Curry Howard correspondence provides us a way to compare two
proofs. The different possibilities of a proof towards the same proposition
gives rise to the intensional structure of type theory (if you admit a
term of some type is a proof). I mainly develop this part in 
\autoref{type-theory} and \autoref{HoTT}. A simple version of the logic
(as you may have observed, this kind of logic is certainly not the
tradtional one, but the intuitionistic logic) isomorphic to the type
theory is described in \autoref{intuitionistic-logic}.

The meta-programming techniques, i.e generating some parts of
the proof from the other, can also be applied to proving. A quick example
is quantification elimination, which provides an algorithm to transfer
each formula to an equivalent and quantification-free one under certain 
circumstances. Haskell turns its language into a category and uses monad 
to represent complicated computation process 
\cite{the-essence-of-functional-programming}. 
In \autoref{intuitionistic-logic}, I give the description of Glivenko's 
theorm and prove it in a monadic style. The concept originated 
in category theory (\autoref{monad}) and is used widely in programming 
\autoref{monad-examples}. Surprisingly, the same style can be applied to 
a proof in intuitionistic logic.
